---
title: "Whole Amazon -- Case Crossover -- All Weeks"
author: "Nicholas J. Arisco"
date: "2024-05-09"
output: html_document
---

# Whole Amazon: Case-Crossover Analysis (All Weeks)

This RMarkdown notebook performs a case-crossover analysis of malaria risk in the Brazilian Amazon. It models relationships between weekly weather metrics and malaria incidence using distributed lag nonlinear models (DLNMs) with generalized nonlinear models (GNMs). Outputs include relative risk (RR) estimates for mean temperature, precipitation, and diurnal variation across 1–3 week exposure windows.

Repository Note:
This script is part of the Malaria & Weather Amazon Analysis repository. Data can be requested through the Brazilian government. We are not at liberty to share.
---

## Packages
```{r}
# Load required libraries
library(mgcv)        # Generalized additive models
install.packages("dlnm") # Install DLNM package for cross-basis functions
library(dlnm)        # Distributed Lag Nonlinear Models
library(splines)     # Spline functions for smooth terms
install.packages("gnm")  # Install GNM package for generalized nonlinear models
library(gnm)         # Generalized nonlinear models
library(data.table)  # Efficient data loading and manipulation
library(dplyr)       # Data wrangling
library(ggplot2)     # Data visualization
```

Load the Data
``` {r}

# Read in malaria and weather dataset
d1m3 <- fread(file = "datafilehere")

# Create year and week variables for temporal stratification
d1m3$year <- year(d1m3$date)
d1m3$week <- week(d1m3$date)  # Epidemiological week
```

Create Cross-Basis Functions
Cross-basis functions are defined for each weather metric at 1-, 2-, and 3-week exposure windows. We use natural cubic splines for variable and lag dimensions.

```{r}

# --- Precipitation (1 week lag) ---
precipitation_knots_1w <- quantile(d1m3$precipitation_1week_sum, c(0.10, 0.75, 0.90))
precipitation_CB1w <- crossbasis(
  d1m3$precipitation_1week_sum, 
  argvar=list(fun="bs", knots=precipitation_knots_1w), 
  lag=0, 
  arglag=list(fun="bs", df=3)
)

# --- Mean Temperature (1 week lag) ---
mean_temp_knots1w <- quantile(d1m3$mean_temp_1week_avg, c(0.10, 0.75, 0.90))
mean_temp_CB1w <- crossbasis(
  d1m3$mean_temp_1week_avg, 
  argvar=list(fun="bs", knots=mean_temp_knots1w), 
  lag=0, 
  arglag=list(fun="bs", df=3)
)

# --- Diurnal Variation (1 week lag) ---
diurnal_knots1w <- quantile(d1m3$diurnal_variation_1week, c(0.10, 0.75, 0.90))
diurnal_CB1w <- crossbasis(
  d1m3$diurnal_variation_1week, 
  argvar=list(fun="bs", knots=diurnal_knots1w), 
  lag=0, 
  arglag=list(fun="bs", df=3)
)

# Repeat for 2-week and 3-week exposure windows
# --- Precipitation (2 weeks) ---
precipitation_knots_2w <- quantile(d1m3$precipitation_2week_sum, c(0.10, 0.75, 0.90))
precipitation_CB2w <- crossbasis(
  d1m3$precipitation_2week_sum, 
  argvar=list(fun="bs", knots=precipitation_knots_2w), 
  lag=0, 
  arglag=list(fun="bs", df=3)
)

# --- Mean Temperature (2 weeks) ---
mean_temp_knots2w <- quantile(d1m3$mean_temp_2week_avg, c(0.10, 0.75, 0.90))
mean_temp_CB2w <- crossbasis(
  d1m3$mean_temp_2week_avg, 
  argvar=list(fun="bs", knots=mean_temp_knots2w), 
  lag=0, 
  arglag=list(fun="bs", df=3)
)

# --- Diurnal Variation (2 weeks) ---
diurnal_knots2w <- quantile(d1m3$diurnal_variation_2week, c(0.10, 0.75, 0.90))
diurnal_CB2w <- crossbasis(
  d1m3$diurnal_variation_2week, 
  argvar=list(fun="bs", knots=diurnal_knots2w), 
  lag=0, 
  arglag=list(fun="bs", df=3)
)

# Repeat for 3-week exposures (precipitation, temperature, diurnal)

```
Run the Model
The model includes all cross-basis functions, extreme weather indicators, and week fixed effects. Individual IDs are eliminated as strata for case-crossover design.

```{r}

model <- gnm(
  formula = case ~ 
    precipitation_CB1w + mean_temp_CB1w + diurnal_CB1w +
    precipitation_CB2w + mean_temp_CB2w + diurnal_CB2w +
    precipitation_CB3w + mean_temp_CB3w + diurnal_CB3w +
    extreme_precipitation + temperature_2m_max + as.factor(week),
  eliminate = as.factor(ID), # Controls for individual-level effects
  family = "quasipoisson",   # Accounts for overdispersion
  data = d1m3
)


# summary(model) # Uncomment for model diagnostics

```

Extract Model Results
Relative risks (RR) for mean temperature, precipitation, and diurnal variation are predicted at each exposure window for plotting.

```{r}
# --- Mean Temperature RR ---
pred.mean_temp1 <- crossreduce(mean_temp_CB1w, model, cen=median(d1m3$mean_temp_1week_avg),
                               value=min(d1m3$mean_temp_1week_avg):max(d1m3$mean_temp_1week_avg))
pred.mean_temp1week <- data.frame(with(pred.mean_temp1, t(rbind(RRfit, RRlow, RRhigh))))
pred.mean_temp1week$lag <- "1 week"

# Repeat for 2 weeks and 3 weeks
pred.mean_temp2 <- crossreduce(mean_temp_CB2w, model, cen=median(d1m3$mean_temp_2week_avg),
                               value=min(d1m3$mean_temp_2week_avg):max(d1m3$mean_temp_2week_avg))
pred.mean_temp2week <- data.frame(with(pred.mean_temp2, t(rbind(RRfit, RRlow, RRhigh))))
pred.mean_temp2week$lag <- "2 weeks"

pred.mean_temp3 <- crossreduce(mean_temp_CB3w, model, cen=median(d1m3$mean_temp_3week_avg),
                               value=min(d1m3$mean_temp_3week_avg):max(d1m3$mean_temp_3week_avg))
pred.mean_temp3week <- data.frame(with(pred.mean_temp3, t(rbind(RRfit, RRlow, RRhigh))))
pred.mean_temp3week$lag <- "3 weeks"

# Combine results
tempRR <- rbind(pred.mean_temp1week, pred.mean_temp2week, pred.mean_temp3week)

# Repeat for Precipitation and Diurnal Variation
# Resulting dataframes: precipitationRR, diurnalRR

```

